hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-28 inst/manuscript/paper_pdf/paper.Rmd

Understanding the timing of key life history events is essential for effective fish sonservation and management. Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on avergae incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings. Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records. However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids. Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes. **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets. The sofware is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for thos unfamiliar with scripting. To illustrate its utility, we present two case studies demonstrating its application in research and management. By expanding assess to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.

>>>>

Understanding the timing of key life history events is essential for effective fish conservation and management. Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings. Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records. However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids. Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes. **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets. The software is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for those unfamiliar with scripting. To illustrate its utility, we present two case studies demonstrating its application in research and management. By expanding assess to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.
